A North Sámi to South Sámi machine translation prototype

### 2014-02-09 paper/paper.tex
La til LREC paper, for tall for  sme-analysator
Litt mer

% Previous work
\cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
% LREC?

@@

Some numbers -- kladd
\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a North Saami-Norwegian dictionary and a South Saami-Norwegian dictionary. The word pairs were manually edited and the work revealed many incorrect pairs because of nob-words with more than one meaning. Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}. The result of this work was 3568 general word pairs and 61 proper nouns.

Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma fremdeles må forklare innholdet ved hjelp av flere ord.

The bilingual dictionary \\
general dictionary\\
general words\\
special domain

\textbf{Problems caused by this model: ex from two terms to one term to two terms}

Example: nob-1,2   sme-x   sma-1,2
example the verbs ‘read, count, say’, nob: \textit{lese, telle, uttale}, sme: \textit{lohkat} can be used about all three, sma: \textit{lohkedh, ryöknehtidh, jiehtedh}

Little nob-sma lexical resources, and total lack of sme-sma, we have to find words by comparing translated texts (nob-sme with nob-sma)

\subsubsection{Lexical selection}

>>>>

% Previous work
\cite{tyers09} \cite{wiechetek10} \cite{trosterud12}
% LREC for numbers for the sme-analyser:
\cite{AntonsenEtalReusing2010}

@@

The bilingual dictionary: Some numbers 
\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

There is no North Saami-South Saami dictionary, so we made one by making word pairs by combining the Norwegian words in a general North Saami-Norwegian dictionary and a general South Saami-Norwegian dictionary. The word pairs were manually edited and the work revealed many incorrect pairs because of Norwegian words with more than one meaning. Example is the Norwegian verb \textit{regne}, which means both 'to rain' and ‘to calculate’, which are two different verbs both in North Saami, \textit{arvit} and \textit{rehkenastit}, and in South Saami \textit{abrodh} and \textit{ryöknedh}. The result of this work was 3568 general word pairs and 61 proper nouns.

The existing Norwegian-South Saami lexical resources are poor, with little terminology from the modern society. South Saami has been used more for giving information to the Saami people especially the last two years, and there are some texts especially about school administration and curriculum on Internet. By comparing Norwegian texts with the translations to North Saami and South Saami, we added 873 special domain word pairs to the bilingual dictionary.

Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami object+verb og adverb+verb. Examples are Norwegian `å trekke lodd' is translated into North Saami \textit{vuorbádit} and North Saami \textit{vuerpiem giesedh}.

Norwegian `å presentere' is in North Saami \textit{ovdanbuktit} but in South Saami \textit{åvtese buektedh}. Often is the reason to this difference that one have utviklet North Saami terms to match the Norwegian terms, and in South Saami one still has to explain the meaning of the term with more words.

These are solved by adding giving the translation as MWE to the morpological analyser.

The pivot model causes some lexical problems. Even if the lexical inndeling often is similar in South Saami and North Saami, as in for the ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}. But in South Saami is no common verb for these three, and the lexical xxx is like in Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.

\subsubsection{Lexical selection}
Because we were tilpasse the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
